### 2015-01-15 fixPoint.tex
change refs to PMR10 and PMR12
begin proofread dynamics section

The study of fix points (or basins of attraction) provides an important understanding of the different behaviors of a BRN (Biological Regulatory Network) \cite{wuensche1998genomic}.  
The fixed point is a stable state of the BRN in wich it is not possible any more to have new changes.
Let $(\Sigma, L, H)$ be a Process Hitting .
 It has been shown that a state $s \in L$ is a fixed point of the Process Hitting if and only if $s$ is has no next state \cite{pauleve2011modelisation} i.e. there is no playable action at this state. In fact a steady state od a Process Hitting network is a set of processes with exactly one process of each sort as well as every process has no hit with the other selected ones (process with self-hit cannot be apart of a stable state).

\subsection{Process Hitting network traduction }
In order to handle a PH biological network, it was necessary first to present it with ASP. To do this we chose the predicates: \texttt{sort}, \texttt{process} and \texttt{action}. Below is an example of PH network written in ASP.

\begin{example*} [Example PH network with ASP]
If we try to present the network of Figure \ref{fig:ph} we will have:
\begin{lstlisting}
sort("a"). sort("b"). sort("z"). %\label{ASPsort}
process("a", 0..1). process("b", 0..1). process("z", 0..2). %\label{ASPprocess}
action("a",0,"b",1,0). action("a",1,"a",1,0). action("b",1,"z",0,2). %\label{actions1}
action("b",0,"z",1,2). action("z",0,"a",0,1). %\label{actions2}
\end{lstlisting}
Line \ref{ASPsort} shows every sort of network with the predicate comes out and in parentheses the name of the sort. In line \ref{ASPprocess} there is a list of processes corresponding to each \texttt{sort}, for example the sort \texttt{"z"} has 3 processes numbered from \texttt{0} to \texttt{2}, this numbering is provided by the $2^{nd}$ parameter of the predicate \texttt{process("z", 0..2)}. Finally we find all network actions that are defined in lines \ref{actions1} and \ref{actions2}, for example the first action \texttt{action("a",0,"b",1,0)} is an action from $a_0$ to $b_1$ to bound so $b$ from $b_1$ into $b_0$.
\end{example*}

\subsection{Search of fix points }
\todo{ Parler du point fixe et ses interrêts. \\
Décrire la nouvelle méthode qui permet de déterminer les états stables d'un réseau }

In fact we have to translate the definition of a stable state in a method developped in ASP. So first we eliminate all processes with self-hit and we save them in the predicate \texttt{shownProcess}:
\begin{lstlisting}
hiddenProcess(A,I) :-  action(A,I,B,J,K), A=B, process(A,I),process(B,J),
                       process(B,K). %\label{hiddenProcess}
shownProcess(A,I) :-  not hiddenProcess(A,I), process(A,I). %\label{shownProcess}
\end{lstlisting}
Then we have to browse this graph and extract all possible combinations of shown processes by choosing a process from each sort.
\begin{lstlisting}
1 { selectProcess(A,I) : showProcess(A,I) } 1 :- sort(A).
\end{lstlisting}
It now remains to check each combination of processes whether it is a fix point. For this we use a special type of ASP rules: a \textbf{constraint}. The idea of constraints is based on the fact that a solution would be eliminated if it does not satisfy the constraint. For our problem a combination is eliminated if there is an action between two of the selected process:
\begin{lstlisting}
:- hit(A,I,B,J), selectProcess(A, I), selectProcess(B, J), A!= B. %\label{contraintFix}
\end{lstlisting}

\begin{example}
Considering the last graphical presentation of a precess hitting \pref{fig:ph}, there are 3 sorts \texttt{"a"}, \texttt{"b"} have 2 levels and \texttt{"z"} has 3, so we can find $2*2*3 = 12$ states (whatever they can be reached or not). If we verify wether there exist fixed points we deduce that we have only one: $<b\_0, z\_2, a\_0>$. It is clear in \pref{fig:ph} that there is no action between each two processes. Besides our new ASP method proof this and returns also the same answer:
\begin{tabbing}
 \texttt{
 Answer 1 : fixProcess(a, 0), fixProcess(b, 0), fixProcess(z, 2)
 }
\end{tabbing}
\end{example}

>>>>

The study of fix points (and, more generally, basins of attraction) provides an important understanding of the different behaviors of a Biological Regulatory Network BRN~\cite{wuensche1998genomic}.
\modMF{%
Indeed, a system will always eventually end in a basin of attraction,
and this may depend on biological switch or other complex phenomena.
}%
A fixed point is a state of the BRN in which it is not possible any more to have new changes;
in other words, it is a basin of attraction that is composed of only one state.

In the following, we consider a Process Hitting $(\Sigma, \PHl, \PHh)$.
It has been shown that a state $s \in L$ is a fixed point of the Process Hitting if and only if $s$ has no next state~\cite{PMR10-TCSB} \ie if there is no playable action in this state.
Therefore, a steady state in a Process Hitting network is a state so that
every process does not hit or is not hit by another process in the same state.
We note that given this result, processes involved in a self-hit cannot be part of a stable state.

\subsection{Process Hitting translation in ASP}
Before writing any piece of ASP,
we first need to present the studied PH network in ASP.
To do this we use the following self-describing predicates:
\texttt{sort}, \texttt{process} and \texttt{action}.
\annotMF{Décrire rapidement le sens de ces prédicats ?}

\begin{example}[Representation of a PH network in ASP]
\label{ex:asp-ph}
The representation of the PH network of Figure \ref{fig:ph} in ASP is the following:
\begin{lstlisting}
sort("a"). sort("b"). sort("z"). %\label{ASPsort}
process("a", 0..1). process("b", 0..1). process("z", 0..2). %\label{ASPprocess}
action("a",0,"b",1,0). action("a",1,"a",1,0). action("b",1,"z",0,2). %\label{actions1}
action("b",0,"z",1,2). action("z",0,"a",0,1). %\label{actions2}
\end{lstlisting}
Line \ref{ASPsort} enumerates every sort of network.
In line \ref{ASPprocess} we create the list of processes corresponding to each sort,
for example the sort \texttt{"z"} has 3 processes numbered from \texttt{0} to \texttt{2};
this specific predicate will in fact expand into the three following predicates:
\texttt{process("z", 0)}, \texttt{process("z", 1)}, \texttt{process("z", 2)}.
Finally, all the actions of the network are defined in lines \ref{actions1} and \ref{actions2};
for example, the first predicate \texttt{action("a",0,"b",1,0)} represents the action
$\PHfrappe{a_1}{b_1}{b_0}$.
\end{example}

\subsection{Search of fix points}
\todo{ Parler du point fixe et ses interrêts. \\
Décrire la nouvelle méthode qui permet de déterminer les états stables d'un réseau }

The enumeration of fix points requires to translate the definition of a stable state
in a set of ASP rules.
The first step consists of eliminating all processes involved in a self-hit;
the others are recorded by the predicate \texttt{shownProcess}:
\begin{lstlisting}
hiddenProcess(A,I) :- action(A,I,B,J,K), A=B, process(A,I),process(B,J),
                      process(B,K). %\label{hiddenProcess}
shownProcess(A,I) :- not hiddenProcess(A,I), process(A,I). %\label{shownProcess}
\end{lstlisting}
Then, we have to browse all remaining processes of this graph
in order to generate all possible states,
that is, all possible combinations of processes by choosing one process from each sort:
\begin{lstlisting}
1 { selectProcess(A,I) : showProcess(A,I) } 1 :- sort(A).
\end{lstlisting}
\modMF{%
The previous line creates an many candidate answer sets as there are possible states
to take into account.
The last step consists of filtering each state that is not a fixed points,
or, in other words, eliminating all candidate answer sets that do not
represent a fixed point.
For this, we use a constraint:
indeed, any solution that satisfy the body of a constraint will be removed from the answer set.
}%
Regarding our problem, a state is eliminated if there exists an action linking two processes:
\begin{lstlisting}
:- hit(A,I,B,J), selectProcess(A, I), selectProcess(B, J), A != B. %\label{contraintFix}
\end{lstlisting}

\begin{example}
The PH model of figure \pref{fig:ph} contains 3 sorts:
$a$ and $b$ have 2 processes and $z$ has 3, therefore, the whole model has $2*2*3 = 12$ states (whether they can be reached or not from a given initial state).
We can check that this model contains only one fixed point: $\PHstate{b_0, z_2, a_0}$.
Indeed, there is no action between each two of the processes contained is this state;
moreover, no other state verifies this property.

If we execute the ASP program detailed below,
alongside with the description of the PS model given in example~\ref{ex:asp-ph},
we obtain one answer set, which matches the expected result:
\begin{lstlisting}[numbers=none]
Answer 1 : fixProcess(a, 0), fixProcess(b, 0), fixProcess(z, 2)
\end{lstlisting}
\end{example}

### 2015-01-15 reachability.tex
begin proofread dynamics section

In this section, we will present firstly how to determine using ASP the possible evolution of a biological network after a finite number of steps. Then we answer to the reachability question; which evolutions that allow the achievement of goals (future active processes) from a known initial state?

\subsection{Future states identification }
From an initial known state, a PH network can evolve into several new states after a few steps. The predicate \texttt{"time (0 .. n)"} sets the number of steps we want to play. For example if the biologist wants to know which states are reachable after 10 stages, it has only to replace \texttt{"n"} by 10, and he will have \texttt{"time (0 .. 10)"}. To compute the future states it is necessary to define an initial state of the network. So for initializing the active processes we add in the ASP script representing the PH network the following rule for each sort. We note that this rules are generated automatically at the time of translating PH network to ASP and by default the level of all active processes is $0$.\\
\begin{lstlisting}
init(activeProcess("a",0)).
\end{lstlisting}
with \texttt{"a"} is the name of the sort and\texttt{"0"} the index of the active process.\\
The dynamic of a network is realised thanks to its actions. So to identify the future states we should start by identifying the playable actions for each state. We recall that an action is playable when both precesses: hitter and taget are active (Definition \ref{def:playableAction}). In ASP the predicate \texttt{"playableAction(A,I,B,J,K,T)"} it is true when the processes \texttt{(A,I)} and \texttt{(B,J)} are active at the step \texttt{T}.
Our approch is to study the asynchronous dynamic so only one action is playable at the same step ( \texttt{T}). That's why between two successive states there will be one change of one sort. We present this by the predicate  \texttt{"activeFromTo(B,J,K,T)"}. It means that in the sort \texttt{B} the active process change from index \texttt{J} to \texttt{K} at time step \texttt{T}.
At the same step we can find many possible changes , that's why the rule (line \ref{e2}) offers a set of all these possible changes thanks to braces.
In fact the rule is encoded with a count atom at its head, which makes it a choice rule. Rule in line \ref{e3} filters any answer with more than 1 change at the same time \texttt{T} to ensure the asynchronous evolution.
\begin{lstlisting}
{activeFromTo(B,J,K,T)} :-  playableAction(A,I,B,J,K,T),
          instate(activeProcess(A,I),T), instate(activeProcess(B,J),T),
          J!=K, time(T). %\label{e2}
:- 2{ activeFromTo(B,J,K,T)}, time(T). %\label{e3}
\end{lstlisting}

>>>>

In this section, firstly present how to determine the possible evolution of a PH model after a finite number of steps with an ASP program.
Then we tackle the reachability question: are there scenarios in a given initial state
that allow to reach a given goal (under the form of a set of processes)?

\subsection{Future states identification}
\reformuler{From an initial known state, a PH network can evolve into several new states after a few steps.}
\modMF{%
Firstly, we focus on the evolutions of the model in a limited number of steps.
}%
The predicate \texttt{time(0..n)} sets the number of steps we want to play.
\modMF{%
The value of \texttt{n} can be arbitrarily chosen.
For example, \texttt{time(0..10)} will be used to compute the 11 first steps,
including the initial state.
In order to specify such an initial state, we add several facts of the following form
to list the processes included in this state:
}%
\begin{lstlisting}
init(activeProcess("a",0)).
\end{lstlisting}
where \texttt{"a"} is the name of the sort and\texttt{"0"} the index of the active process.
The dynamics of a network is described by its actions,
identifying the future states requires to first identify the playable actions for each state.
We recall that an action is playable in a state when both its hitter and target are active in this state (see Definition \ref{def:playableAction}).
Therefore, we define an ASP predicate \texttt{playableAction(A,I,B,J,K,T)} that is true
when the processes \texttt{(A,I)} and \texttt{(B,J)} are active at step \texttt{T}.
\annotMF{Peut-être écrire : $\texttt{A}_\texttt{I}$ et $\texttt{B}_\texttt{J}$}
It is also needed to enforce the strictly asynchronous dynamic
which state that exactly one process can change between two steps.
We thus represent the change of the active process of a sort
by the predicate \texttt{"activeFromTo(B,J,K,T)"}
which means that in sort \texttt{B}, the active process changes from index \texttt{J} to \texttt{K} between steps \texttt{T} and \texttt{T+1}.
\annotMF{Peut-être écrire : $\texttt{B}_\texttt{J}$ et $\texttt{B}_\texttt{K}$}
\modMF{%
The cardinality rule of line \ref{e2} then
created as many answer sets as there are possible evolutions from the current step,
thus recreating all possible branchings in the possible evolutions of the model.
This allows to filter all scenarios where two actions have been played between
two steps, by using the constraint of \ref{e3}.
Thus, the remaining scenarios of the answer set all strictly follow
the asynchronous dynamics of the PH.
}%
\begin{lstlisting}
{activeFromTo(B,J,K,T)} :-  playableAction(A,I,B,J,K,T),
          instate(activeProcess(A,I),T), instate(activeProcess(B,J),T),
          J!=K, time(T). %\label{e2}
:- 2{ activeFromTo(B,J,K,T)}, time(T). %\label{e3}
\end{lstlisting}
\annotMF{N'est-il pas possible de réunir ces deux règles avec des bornes 1--1 de la façon suivante : 1 \{activeFromTo(B,J,K,T)\} 1 $\leftarrow$ ... ?}
